\thetitle: Technical Report

### 2023-04-02 metatheory-condensed.tex
add a forward reference as suggested by shepherd
tweak
TR

For accessibility of the main text, \cref{sec:metatheory2} presents \TT using
named variables and meta-level substitution.
For clarity of proof details, a technical appendix supplementing
\cref{sec:metatheory2} instead uses \citeauthor{debruijn1964} indices and explicit substitution \cite{abadi1989subst}.
%
We acknowledge that the presentation is dense for an audience
without intimate knowledge of MLTT, so we summarize the salient points first.

@@

\cref{fig:mltt-selected} presents the syntax and selected typing rules of MLTT %(predicative) MLTT
(and \cref{fig:fmltt-selected} \TT).
Dependent function types \nTyPi{x}{A}{B},
dependent pair types \nTySigma{x}{A}{B}, and
identity types \TyId{t_1}{t_2} are standard.
So are their introduction and elimination forms.
We use based path induction $\Jeq{\cdot}{r}$
as the elimination principle for 
a term $r$ of identity types~\citepalias{program2013homotopy}. 
Capture-avoiding substitution is notated $\bullet[\bullet/x]$.
We use $x$ and $\mathit{self}$ to denote variables.
A singleton type $\TyS{t}$ helps expose the definition of a term~$t$
in its type (rule \ruleref{tm/s})~\cite{aspinall1995singleton,stone2000}.
Definitional equalities have the forms %(for each judgement)
\goodCtx{\Gm_1 \equiv \Gm_2}{},
\goodType{\Gm}{T_1 \equiv T_2}{},
\goodTerm{\Gm}{t_1 \equiv t_2}{T},
etc.
Following \citet{altkap2016}, we regard our (intrinsically typed) syntax
as being quotiented by these equalities.
Quotienting facilitates coercion along equalities---given
\goodCtx{\Gm_1 \equiv \Gm_2}{} and
\goodType{\Gm_1}{T_1 \equiv T_2}{},
the derivation of \goodTerm{\Gm_1}{t}{T_1} is considered definitionally equal to
a derivation of \goodTerm{\Gm_2}{t}{T_2}. %\YZ{Is this accurate??}\EDJreply{Yes. In QIIT model, it looks like `t : Tm G1 T1' and  `t : Tm G2 T2'. In extenional Agda, because ``G1 = G2'', ``T1 = T2'', thus the two types `Tm G2 T2 = Tm G1 T1'. So they are the same derivation in QIIT. In our setup, we have the same idea, because after quotient, G1 refers to same object as G2, so the derivation is the same.}
%\YZ{should there be a convertibility rule?}\EDJreply{in this style, the quotient by judgemental equality is applied, so A is equal to B even in meta-logic, which means t : A is the same as t : B. Maybe this is something require emphasis? Because it is usually handled by LF (like QIIT) but now we are using bare hands so.... maybe we need to emphasize this}
%For each derivation (of each judgment)\YZ{I don't get what 'each derivation' or 'each judgment' means}\EDJreply{(G |- t : T), which looks like (t : Tm G T) in QIIT/Agda is the derivation; so basically the term is the derivation (or proof), the types (Tm G T) is the judgement. It applies to each kind of judgement, for example, (G |- ), which looks like (G : Con), (G |-) is the derivation and (_ |- )is the judgement \\ 
%Note that without referring to QIIT/Agda, the definition of judgement and derivation is still clear -- (G |- _ : T) is a judgement and a concrete proof/derivation we get for t like (G |- t : T) is the derivation. Similar for ( _ |- ) vs. (G |-) and (G |- _ ) vs (G |- T). \\ 
%I think in intrinsic typing, knowing t and knowing (G |- t : T) is exactly the same (i.e. there are no two different derivations lead to the same t; for each t, there is only one derivation for it). This is not true for extrinsic typing. \\
%If this prequisite knowledge is too involved, feel free to delete this sentence, I think I just want to say there are equality/quotient everywhere between all kinds of derivations(proof?data?) -- equality between terms, types, substitutions, contexts, linkage signatures ...\\
%I think this is something standard from Martin Lof, since Martin Lof is the first guy talk about judgement.
%}

@@

In \ruleref{l/add} and \ruleref{lsig/add},
the third premise $\goodTerm{\Gm,x : \TyTkg{\lsig}}{\seal}{A}$ is
responsible for hiding W-type signatures.
Here, $\TyTkg{\lsig}$ packages $\lsig$---recall that $\lsig$ contains the
(self-parameterized) types of all the fields preceding the current
field~$t$---into a dependent tuple type (\ruleref{tyeq/pk/add}).
The term $\seal$ turns a tuple of type $\TyTkg{\lsig}$ into a new tuple of type~$A$
that hides W-type signatures behind $\cU$, if necessary.\footnote{%
In the presence of hiding, it is important that the context type~$A$ be a
dependent tuple type rather than a linkage type; we expand on this point in an
appendix.}

@@

The definitive version containing all the rules in \TT is stated in
\ifreport \cref{sec:complete-fmltt}.
\else a technical appendix.
\fi
This definitive version uses de Bruijn indices and explicit substitution.
An effort to mechanize the formalization is underway. It follows \citet{altkap2016}
in using \ac{QIIT} as 
the logical framework to state intrinsically typed syntax (and also
in using Agda's \lsti{REWRITE} pragma to replace the limited uses of equality
reflection, as Agda does not support QIITs natively).
%in the extensional meta\-logic.

>>>>

For accessibility of the main text, \cref{sec:metatheory2} presents \TT using
named variables and meta-level substitution.
For clarity of proof details,
\ifreport
\cref{sec:complete-fmltt}
\else
a technical appendix \cite{fpop-tr}
\fi
supplementing \cref{sec:metatheory2}
instead uses \citeauthor{debruijn1964} indices and explicit substitution \cite{abadi1989subst}.
%
We acknowledge that the presentation is dense for an audience
without intimate knowledge of MLTT, so we summarize the salient points first.

@@

\cref{fig:mltt-selected} presents the syntax and selected typing rules of MLTT %(predicative) MLTT
(and \cref{fig:fmltt-selected} \TT).
Dependent function types \nTyPi{x}{A}{B},
dependent pair types \nTySigma{x}{A}{B}, and
identity types \TyId{t_1}{t_2} are standard.
So are their introduction and elimination forms.
We use based path induction $\Jeq{\cdot}{t}$
as the elimination principle for 
a term~$t$ of an identity type~\citepalias{program2013homotopy}. 
Capture-avoiding substitution is notated $\bullet[\bullet/x]$.
We use $x$ and $\mathit{self}$ to denote variables.
A singleton type $\TyS{t}$ helps expose the definition of a term~$t$
in its type (rule \ruleref{tm/s})~\cite{aspinall1995singleton,stone2000}.
Definitional equalities have the forms %(for each judgement)
\goodCtx{\Gm_1 \equiv \Gm_2}{},
\goodType{\Gm}{T_1 \equiv T_2}{},
\goodTerm{\Gm}{t_1 \equiv t_2}{T},
etc.
Following \citet{altkap2016}, we regard our (intrinsically typed) syntax
as being quotiented by these equalities.
Quotienting facilitates coercion along equalities---given
\goodCtx{\Gm_1 \equiv \Gm_2}{} and
\goodType{\Gm_1}{T_1 \equiv T_2}{},
the derivation of \goodTerm{\Gm_1}{t}{T_1} is considered definitionally equal to
a derivation of \goodTerm{\Gm_2}{t}{T_2}. %\YZ{Is this accurate??}\EDJreply{Yes. In QIIT model, it looks like `t : Tm G1 T1' and  `t : Tm G2 T2'. In extenional Agda, because ``G1 = G2'', ``T1 = T2'', thus the two types `Tm G2 T2 = Tm G1 T1'. So they are the same derivation in QIIT. In our setup, we have the same idea, because after quotient, G1 refers to same object as G2, so the derivation is the same.}
%\YZ{should there be a convertibility rule?}\EDJreply{in this style, the quotient by judgemental equality is applied, so A is equal to B even in meta-logic, which means t : A is the same as t : B. Maybe this is something require emphasis? Because it is usually handled by LF (like QIIT) but now we are using bare hands so.... maybe we need to emphasize this}
%For each derivation (of each judgment)\YZ{I don't get what 'each derivation' or 'each judgment' means}\EDJreply{(G |- t : T), which looks like (t : Tm G T) in QIIT/Agda is the derivation; so basically the term is the derivation (or proof), the types (Tm G T) is the judgement. It applies to each kind of judgement, for example, (G |- ), which looks like (G : Con), (G |-) is the derivation and (_ |- )is the judgement \\ 
%Note that without referring to QIIT/Agda, the definition of judgement and derivation is still clear -- (G |- _ : T) is a judgement and a concrete proof/derivation we get for t like (G |- t : T) is the derivation. Similar for ( _ |- ) vs. (G |-) and (G |- _ ) vs (G |- T). \\ 
%I think in intrinsic typing, knowing t and knowing (G |- t : T) is exactly the same (i.e. there are no two different derivations lead to the same t; for each t, there is only one derivation for it). This is not true for extrinsic typing. \\
%If this prequisite knowledge is too involved, feel free to delete this sentence, I think I just want to say there are equality/quotient everywhere between all kinds of derivations(proof?data?) -- equality between terms, types, substitutions, contexts, linkage signatures ...\\
%I think this is something standard from Martin Lof, since Martin Lof is the first guy talk about judgement.
%}

@@

In \ruleref{l/add} and \ruleref{lsig/add},
the third premise $\goodTerm{\Gm,x : \TyTkg{\lsig}}{\seal}{A}$ is
responsible for hiding W-type signatures.
Here, $\TyTkg{\lsig}$ packages $\lsig$---recall that $\lsig$ contains the
(self-parameterized) types of all the fields preceding the current
field~$t$---into a dependent tuple type (\ruleref{tyeq/pk/add}).
The term $\seal$ turns a tuple of type $\TyTkg{\lsig}$ into a new tuple of type~$A$
that hides W-type signatures behind $\cU$, if necessary.\footnote{%
In the presence of hiding, it is important that the context type~$A$ be a
dependent tuple type rather than a linkage type; we expand on this point in
\ifreport
\cref{sec:complete-fmltt}.
\else
an appendix.
\fi
}

@@

The definitive version containing all the rules in \TT is stated in
\ifreport \cref{sec:complete-fmltt}.
\else a technical appendix \cite{fpop-tr}.
\fi
This definitive version uses de Bruijn indices and explicit substitution.
An effort to mechanize the formalization is underway. It follows \citet{altkap2016}
in using \ac{QIIT} as 
the logical framework to state intrinsically typed syntax (and also
in using Agda's \lsti{REWRITE} pragma to replace the limited uses of equality
reflection, as Agda does not support QIITs natively).
%in the extensional meta\-logic.
